---
title: "Revealing the enduring effect of social class among researchers´s scientific productivity: An analysis from Chile"
author: 
  - name: "Roxana Chiappa"
    affiliation: "Universidad de Tarapacá"
    email: ""
  - name: "Paula Clasing"
    affiliation: ""
    email: ""
  - name: "Carlos Anríquez"
    email: ""
date: today
abstract: |
 
keywords: ["geographical disparities", "interseccionality", "PHD education", "academic inequalities"]
lang: es
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = 'center',
  fig.width = 8,
  fig.height = 6)

# Cargar librerías necesarias
pacman::p_load(
  dplyr, ggplot2, tidyverse, tibble, scales, purrr, 
  kableExtra, sessioninfo, sjPlot, stringr, knitr, patchwork)

# Tema personalizado para gráficos
theme_academic <- theme_minimal() +
  theme(
    text = element_text(family = "serif", size = 12),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(face = "bold"),
    legend.position = "bottom",
    panel.grid.minor = element_blank())

theme_set(theme_academic)

```

# Introduction

One of the most robust findings in the higher education literature is that as access expands, actors—both institutions and individuals—occupying privileged positions develop new strategies of differentiation to maintain their relative advantage (xxxxx). This premise lies at the core of Bourdieu and Passeron’s theory of social and cultural reproduction (1977), as well as of sociological approaches concerned with explaining the persistence and, in some cases, the intensification of inequalities across educational trajectories. Theories such as Maximum Maintained Inequality (Raftery & Hout, 1993), the Matthew Effect (Merton, 1968), and related formulations of persistent inequality share a common assumption grounded in cumulative advantage. From this perspective, early access to valued forms of capital—economic, cultural, social, or institutional—generates self-reinforcing processes through which advantages compound over time, systematically positioning already privileged individuals and institutions to secure further resources, recognition, and opportunities.
In the process of expansion of doctoral education, 
This dynamic is especially evident in scholarship examining the role of social class of origin in shaping access to prestigious universities in academic careers (Zhang, xxxx). Individuals from upper social class backgrounds are consistently more likely to enroll in elite institutions, where they gain access to networks, resources, and forms of academic socialization that later function as key assets in scientific careers. These forms of social and cultural capital facilitate subsequent transitions, particularly job placement and career advancement within similarly prestigious academic institutions (xxxxx). Being hired to a highly prestigious university is also associated with greater academic productivity.  
A similar logic of linear accumulation has also structured much of the literature on gender inequalities in scientific careers. Despite women’s increased access to higher education and, in many contexts, their equal or superior academic performance compared to men, their representation declines markedly at higher levels of the academic hierarchy. This pattern is often captured through the metaphor of the “leaky pipeline,” which conceptualizes gender inequality as the cumulative result of successive barriers encountered by women as they advance through an academic system historically structured around androcentric norms.
In a highly centralized academic system such as Chile’s, regional academic careers are shaped by distinct incentive structures, organizational pressures, and evaluative regimes. For scholars outside the metropolitan core, scientific productivity may function less as a cumulative reward of advantage and more as a strategy of institutional survival and professional legitimacy within contexts marked by lower symbolic capital and greater precarity.
These theoretical frameworks have been central to understanding the mechanisms that shape access and early-stage selection—particularly for individuals from lower social class backgrounds and for women. Nonetheless, the assumption of linear cumulative advantage becomes less sufficient when the analytical focus shifts to trajectories among those who have already crossed key institutional thresholds.
 Once individuals have entered doctoral education, secured competitive fellowships, or obtained positions within the academic field, the population under observation is no longer open or broadly stratified, but highly selected, where institutional arrangements and agential mechanism had a role informing those who are able to access. 
Rather than producing a straightforward continuation of cumulative advantage, inequalities at these later stages tend to reconfigure.  Structural hierarchies such as gender, social class, or region of origin continue to matter, but their effects may vary in magnitude or even direction depending on the specific stage of the career and the outcomes under consideration. This reconfiguration reflects shifts in what counts as valuable capital within the academic field and in how different actors are positioned to convert their resources into scientific recognition, stability, or productivity. Consequently, patterns of advantage among those who “make it” cannot be fully captured by linear models of accumulation, but instead require an analytical framework attentive to stage-specific mechanisms of stratification and to the contextual conditions under which inequalities are reproduced, transformed, or partially offset.
In this paper, we propose a longitudinal analysis of the effect of three structural dimensions of inequality that shape educational opportunities: gender, class, and regional differences, and explore how they affect key moments of the scientific careers: The selection of the undergraduate university, the decision of studying a doctoral degree, and its publication records after the five years since they obtain the fellowship. 
Examining a cohort of scholars selected to pursue doctoral training in Chile and abroad through competitive fellowship programs, our findings reveal a counterintuitive pattern: researchers originating from regions outside the metropolitan center exhibit higher levels of scientific productivity, measured by the number of peer-reviewed publications, than their counterparts from the metropolitan area. This pattern does not suggest the absence of inequality, nor does it imply a reversal of structural hierarchies. Rather, it points to the need for a situated understanding of how inequalities are produced, negotiated, and sometimes reconfigured within specific institutional and spatial contexts.

Analysis of scholarship policies from a perspective of inequiality
Over the past two decades, the global expansion of higher education has extended into the realm of doctoral training, widely regarded as the highest academic degree. While the United States continues to host the largest number of doctoral enrollments (Grant et al., 2021; OECD, n.d.), several middle-income countries in the so-called Global South have significantly expanded their doctoral education systems, often through state-funded fellowship programs (for Brazil, see Chiappa & Finardi, 2021; for Mexico, see Murillo, 2023; for Chile, see Chiappa & Labraña, 2023; for multiple cases, see Celis et al., 2025).
These initiatives— implemented in countries such as Chile, Brazil, and Mexico—have been typically framed by meritocratic discourses that claim to reward the most talented individuals (see SEICHTI, n.d., CNIC, 2007; McManus and Nobre, 2017). They offer competitive funding to pursue doctoral studies either domestically or at internationally prestigious institutions, with the broader aim of boosting national scientific capacity and global competitiveness.
However, the implementation of such initiatives is exposed to at least two levels of stratification. On the one


Las becas ---> y la politica 
Inequalities in scientific careers

Research on scientific careers has long emphasized the cumulative and self-reinforcing nature of inequality. Classic formulations of cumulative advantage and the Matthew effect posit that early access to resources, institutional prestige, and recognition generates compounding processes through which advantages accrue over time, shaping long-term academic productivity and career outcomes. This framework has been highly influential in explaining persistent gender, class, and institutional inequalities in scientific fields, particularly within highly stratified academic systems.
Much of this literature relies on an implicitly linear understanding of inequality reproduction, assuming that disadvantage uniformly constrains performance and that proximity to centers of power and resources consistently translates into higher productivity. This perspective has been especially prominent in research on the underrepresentation of women across academic trajectories in different national contexts, revealing a series of sorting and filtering mechanisms that systematically limit women’s access to senior academic positions as careers progress.
However, this linear and accumulative approach becomes less suitable for examining stratification dynamics among those who have already survived and overcome these filtering mechanisms. Once access barriers have been crossed, academic trajectories are shaped not only by accumulated advantage, but also by context-specific institutional arrangements, differentiated incentive structures, and the agential strategies scholars deploy under unequal conditions. By prioritizing cumulative disadvantage as a universal explanatory mechanism, much of the existing literature underestimates the contextual and relational processes through which inequalities are reworked within academic careers.

# Conceptual framework: stratification on sciences.

## Geographical disparities in academic trayectories

## Socioeconomic status

## Disciplinary differences in scientific production

## Gender and scientific production

## Scolarship selection process

# Method

## Data and sample

The database is a longitudinal reconstruction of the academic trajectories of three cohorts of Chilean researchers (N = 2436) who received a doctoral scholarship between 2014 and 2016. It includes information on the socioeconomic level of the high school they attended, the region of the institution, the accreditation level of their undergraduate program, the prestige of the doctoral university, the postgraduate area, and the number of publications produced up to six years after the scholarship was awarded.


## Variables

Table 1 shows the variables used in the analysis. These summarize the researchers' trajectory from secondary education to scientific productivity after the doctorate. The analysis employed three sociodemographic variables: sex, socioeconomic status, and region. To measure socioeconomic status, the socioeconomic classification of each researcher's high school served as a proxy, while geographic origin was determined by the educational institution's region.

Undergraduate prestige was measured through the accreditation level of the researcher's university as determined by the National Education Commission. For doctoral prestige, we calculated the average between the scores in the Shanghai and QS international rankings, expressed as a Z-score. Finally, scientific productivity was measured through the number of publications, standardized based on the average publications in the researcher's area.

```{r cargar bb.dd, echo= FALSE}

df_senderos <- read_rds("output/df_senderos.rds")

```

```{r tabla-variables, echo=FALSE}

# formato long
tabla_cat_procesada <- df_senderos %>%
  dplyr::select(SEXO_REGISTRADO, GSE_ESTABLECIMIENTO, REGIONES_3, 
                ACREDITACION_PREGRADO, tipuban, CHILEDOCTORADO) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Valor") %>%
  drop_na() %>%
  count(Variable, Valor) %>%
  group_by(Variable) %>%
  mutate(Porcentaje = round(n / sum(n) * 100, 1),
         Estadistico = paste0(n, " (", Porcentaje, "%)")) %>%
  ungroup() %>%
  dplyr::select(Variable, Valor, Estadistico)

# formato long var numericas 
tabla_num_procesada <- df_senderos %>%
  dplyr::select(Average, z_scopus_despues) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Valor_num") %>%
  drop_na() %>%
  group_by(Variable) %>%
  summarise(
    Min = round(min(Valor_num, na.rm = TRUE), 2),
    Max = round(max(Valor_num, na.rm = TRUE), 2),
    Mean = round(mean(Valor_num, na.rm = TRUE), 2),
    SD = round(sd(Valor_num, na.rm = TRUE), 2)) %>%
  pivot_longer(cols = c(Min, Max, Mean, SD), 
               names_to = "Valor", 
               values_to = "Estadistico") %>%
  mutate(Estadistico = as.character(Estadistico)) %>%
  dplyr::select(Variable, Valor, Estadistico)

# Combinar ambas tablas
tabla_combinada <- bind_rows(
  tabla_cat_procesada,
  tabla_num_procesada) %>%
  mutate(Variable = case_when(
         Variable == "SEXO_REGISTRADO" ~ "Gender",
         Variable == "GSE_ESTABLECIMIENTO" ~ "Socio-economic Status",
         Variable == "REGIONES_3" ~ "Region",
         Variable == "ACREDITACION_PREGRADO" ~ "Undergraduate Prestige",
         Variable == "tipuban" ~ "Publication before PhD",
         Variable == "Average" ~ "PhD Prestige (Z)",
         Variable == "CHILEDOCTORADO" ~ "National PhD",
         Variable == "z_scopus_despues" ~ "Publications after PhD (Z)",
         TRUE ~ Variable),
    Valor = case_when(
      Valor == "Min" ~ "Min.",
      Valor == "Max" ~ "Max.",
      Valor == "Mean" ~ "Mean",
      Valor == "SD" ~ "SD",
      TRUE ~ Valor),
    orden = case_when(
      Variable == "Gender" ~ 1,
      Variable == "Socio-economic Status" ~ 2,
      Variable == "Region" ~ 3,
      Variable == "Undergraduate Prestige" ~ 4,
      Variable == "Publication before PhD" ~ 5,
      Variable == "PhD Prestige (Z)" ~ 6,
      Variable == "National PhD" ~ 7,
      Variable == "Publications after PhD (Z)" ~ 8,
      TRUE ~ 99)) %>%  arrange(orden, Valor)

# tabla combinada

tabla_combinada <- tabla_combinada %>%
  mutate(fila = row_number())

indices_grupos <- tabla_combinada %>%
  group_by(Variable) %>%
  summarise(
    inicio = min(fila),
    fin = max(fila),
    .groups = "drop")

# formato kable
tabla_kable <- tabla_combinada %>%
  dplyr::select(Valor, Estadistico) %>%
  knitr::kable(booktabs = TRUE,
               col.names = c("Category", "Statistic"),
               caption = "Table 1. Descriptive Statistics",
               align = c("l", "r")) %>%
  kableExtra::kable_styling(
    latex_options = c("hold_position", "striped"),
    font_size = 11,
    full_width = FALSE)

# ordenar tabla
for(i in 1:nrow(indices_grupos)) {
  tabla_kable <- tabla_kable %>%
    kableExtra::pack_rows(
      indices_grupos$Variable[i],
      indices_grupos$inicio[i],
      indices_grupos$fin[i],
      bold = TRUE,
      latex_gap_space = "0.5em")}

tabla_kable

```

```{r outliers-productividad, echo=FALSE}

p2 <- ggplot(df_senderos %>% filter(z_scopus_despues > 3), 
            aes(x = DISCIPLINA_DOCTORADO, y = z_scopus_despues, 
                color = DISCIPLINA_DOCTORADO)) +
  geom_jitter(width = 0.2, alpha = 0.6, size = 2) +
  labs(title = "Outliers (Z > 3)", x = "", y = "Valor Z") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_blank())

p1 <- ggplot(df_senderos, aes(x = DISCIPLINA_DOCTORADO, y = z_scopus_despues, 
                              fill = DISCIPLINA_DOCTORADO)) +
  geom_violin(alpha = 1, color = NA) +
  stat_summary(fun = median, geom = "point", size = 1, color = "darkgrey") +
  coord_cartesian(ylim = c(-2, 3)) +
  labs(title = "", x = "", y = "Valor Z (Scopus post-doctoral)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 9))


p2 / p1

```

## Analytic Strategy

In accordance with the research objectives and the hypotheses proposed, we opted to use path analysis. This technique uses regression models to represent theories of causal relationships. In statistical terms, it consists of a series of ordered regressions within the chosen analytical "path". The main advantage of this method is that it allows to make explicit the relationships between the selected variables.  Additionally, it enables the decomposition of direct and indirect effects (Agresti, 2018, p. 506).

# Finding

In accordance with the hypotheses proposed, social class has a substantial effect on the quality of undergraduate education to which researchers gain access. Compared to the high socioeconomic group, the upper-middle and middle groups decrease the quality of undergraduate education (-0.388 and -0.366, respectively), while the low group has a strong negative effect on undergraduate quality (-0.741). Moreover, the region of origin shows a statistically significant effect. Compared to other regions, completing secondary education in Santiago, Valparaíso, and Concepción has a positive and moderate effect on undergraduate quality (0.367). Finally, gender does not present a statistically significant effect.

Regarding prior publications, neither socioeconomic group nor undergraduate quality have a statistically significant effect. Being female and coming from the metropolitan, Valparaíso, and Concepción regions have a small negative effect (-0.038 and -0.041).

The effects on doctoral prestige are also revealing. Compared to the high socioeconomic group, the other groups decrease doctoral prestige. The upper-middle group has a small negative effect (-0.127), the middle group has a moderate effect (-0.205), and the low group has a moderate to strong effect (-0.331).

Concerning region, coming from the main urban centers increases doctoral prestige (0.134). Undergraduate accreditation has a small but positive effect (0.091). In line with expectations, completing the doctorate in Chile has a negative effect on doctoral prestige (-1.073). Gender and the number of prior publications do not have statistically significant effects on doctoral prestige.

Finally, with respect to publications after the doctorate, socioeconomic group, region, and gender have important effects on scientific productivity. Compared to men, being female presents a moderate negative effect (-0.120). Similarly, compared to the high socioeconomic group, the middle group has a statistically significant negative effect (-0.106) on scientific productivity. However, the low group does not present a statistically significant effect.

On the other hand, coming from the main urban centers has a small but negative effect on scientific productivity (-0.141). Additionally, doctoral prestige presents a small negative effect (-0.053), as does having completed the doctorate in Chile (-0.131). Undergraduate accreditation does not have a statistically significant effect.


# Conclusions

## Referencias

::: {#refs}


:::
